The \dealii Library, Version 9.0

### 2018-04-25 9.0/paper.tex
Add section on automatic differentiation

The major changes of this release are:
\begin{itemize}
\item Improved support for curved geometries;
\item Support for particle-in-cell methods;
\item Dedicated support for automatic and symbolic differentiation;
\item Interfaces to more external libraries and programs;
\item C++11 is now both required and used;
\item Support for GPU computations;
\item Support for face integrals and significant improvements of the matrix-free framework;
\end{itemize}
These will all be discussed in more detail in the
following section. In addition, this release contains the following changes:
\begin{itemize}
\item \dealii has made extensive use of both the Clang-Tidy \cite{clang-tidy}
  and Coverity Scan \cite{coverity} static analysis tools for detecting bugs
  and other issues in the code. For example, around 260 issues were detected and
  fixed using the latter tool.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Dedicated support for automatic and symbolic differentiation}

\marginpar{J-P: Please edit as appropriate}
Automatic differentiation is often used to automatically derive
residuals from the stored energy functional, and to derive Jacobian
matrices from residual vectors for simulations that use complicated
material models. Examples can be found widely for nonlinear solid
mechanics as well as for nonlinear viscosity models in fluid flow.

\dealii has had a tutorial program (step-33) since 2007 that
demonstrates this technique based on the Trilinos Sacado package, but
the functionality was not available pervasively throughout
deal.II. This has changed with release 9.0 where support for
differentiation has been collected in one module that provides both
automatic and symbolic differentiation. Currently, this module relies
on either the ADOL-C and Sacado libraries, through a unified
interface. In practice, this support means that ADOL-C and Sacado data
types can be used in the \texttt{FEValues}, \texttt{FEValuesViews},
and related classes that are generally used to assemble linear systems
and right hand sides.

@@

\dealii can interface with many other libraries:
\begin{multicols}{3}
\begin{itemize}
\item ADOL-C \cite{adol-c}
\item ARPACK \cite{arpack}
\item Assimp \cite{assimp}
\item BLAS and LAPACK \cite{lapack}
\item Gmsh \cite{geuzaine2009gmsh}
\item GSL \cite{gsl2016}
\item HDF5 \cite{hdf5}
\item METIS \cite{karypis1998fast}
\item MUMPS \cite{ADE00,MUMPS:1,MUMPS:2,mumps-web-page}
\item muparser \cite{muparser-web-page}
\item nanoflann \cite{nanoflann}
\item NetCDF \cite{rew1990netcdf}
\item OpenCASCADE \cite{opencascade-web-page}
\item p4est \cite{p4est}
\item PETSc \cite{petsc-user-ref,petsc-web-page}
\item ROL \cite{ridzal2014rapid}
\item ScaLAPACK \cite{slug}
\item SLEPc \cite{Hernandez:2005:SSF}
\item SUNDIALS \cite{sundials}
\item TBB \cite{Rei07}
\item Trilinos \cite{trilinos,trilinos-web-page}
\item UMFPACK \cite{umfpack}
\end{itemize}
\end{multicols}
Please consider citing the appropriate references if you use interfaces to these
libraries.

>>>>

The major changes of this release are:
\begin{itemize}
\item Improved support for curved geometries;
\item Support for particle-in-cell methods;
\item Dedicated support for automatic differentiation;
\item Interfaces to more external libraries and programs;
\item C++11 is now both required and used;
\item Support for GPU computations;
\item Support for face integrals and significant improvements of the matrix-free framework;
\end{itemize}
These will all be discussed in more detail in the
following section. In addition, this release contains the following changes:
\begin{itemize}
\item \dealii has made extensive use of both the Clang-Tidy \cite{clang-tidy}
  and Coverity Scan \cite{coverity} static analysis tools for detecting bugs
  and other issues in the code. For example, around 260 issues were detected and
  fixed using the latter tool.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Dedicated support for automatic differentiation}

Automatic differentiation (AD) is often used to automatically derive
residuals and their linearization from a stored energy functional, and to
derive Jacobian matrices from residual vectors for simulations that use
complicated material models. Examples of its application can be found widely
within nonlinear solid mechanics, coupled multiphysics problems, as well as
for nonlinear viscosity models in fluid flow.

\dealii has had a tutorial program (step-33) since 2007 that
demonstrates this technique based on the Trilinos Sacado \cite{Bartlett2006a} package, but
the functionality was not available pervasively throughout
deal.II. 
This has changed with release 9.0 where support is given for differentiation 
using a selection of ``white-listed'' libraries (namely ADOL-C \citep{Griewank1996a} 
and Sacado) and a subset of their supported number types. Currently, we offer support for
\begin{itemize}
\item ADOL-C taped (n-differentiable),
\item ADOL-C tapeless (once differentiable),
\item Sacado dynamic forward (once differentiable),
\item Sacado reverse (once differentiable),
\item Sacado nested dynamic forward (twice differentiable), and
\item Sacado nested reverse and dynamic forward (twice differentiable)
\end{itemize}
numbers. In practice, this support means that these ADOL-C and Sacado data
types can be used in the \texttt{FEValues}, \texttt{FEValuesViews},
\texttt{Tensor}, \texttt{SymmetricTensor},
and related classes that are generally used to assemble linear systems
and right hand sides.
Given the updated capabilities of the library, there is now a dedicated module 
that presents the AD compatibility and capabilities of the \dealii libraries. 
Furthermore, the use of Sacado is now  demonstrated in a much more simplified 
and transparent manner in an modernized version of an existing ``code gallery'' 
example \cite{Pelteret2016a}.

To date it remains necessary for the user  to manage the initialization of AD 
independent variables and the resultant calls  to the AD dependent variables 
in order to initiate the computation of derivatives.
In the next release we expect to provide a unified interface to these
AD libraries, which will hide these library-dependent implementational details 
and facilitate switching between the supported libraries and AD number types
based on the user's requirements.

@@

\dealii can interface with many other libraries:
\begin{multicols}{3}
\begin{itemize}
\item ADOL-C \cite{Griewank1996a,adol-c}
\item ARPACK \cite{arpack}
\item Assimp \cite{assimp}
\item BLAS and LAPACK \cite{lapack}
\item Gmsh \cite{geuzaine2009gmsh}
\item GSL \cite{gsl2016}
\item HDF5 \cite{hdf5}
\item METIS \cite{karypis1998fast}
\item MUMPS \cite{ADE00,MUMPS:1,MUMPS:2,mumps-web-page}
\item muparser \cite{muparser-web-page}
\item nanoflann \cite{nanoflann}
\item NetCDF \cite{rew1990netcdf}
\item OpenCASCADE \cite{opencascade-web-page}
\item p4est \cite{p4est}
\item PETSc \cite{petsc-user-ref,petsc-web-page}
\item ROL \cite{ridzal2014rapid}
\item ScaLAPACK \cite{slug}
\item SLEPc \cite{Hernandez:2005:SSF}
\item SUNDIALS \cite{sundials}
\item TBB \cite{Rei07}
\item Trilinos \cite{trilinos,trilinos-web-page}
\item UMFPACK \cite{umfpack}
\end{itemize}
\end{multicols}
Please consider citing the appropriate references if you use interfaces to these
libraries.
